---
title: "Anatomy of a Winsteps run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anatomy of a Winsteps run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Winsteps is not a library you call. It is a program that reads a set of plain
text files from a directory, writes another set of files back, and exits. Every
function in this package exists to produce one of those files or to read one of
them, so the fastest way to understand the package is to look at the files.

Everything below runs without Winsteps installed. The input files are generated
here; the output files are real Winsteps output shipped with the package.

```{r setup}
library(winstepsR)
```

## The data going in

Responses arrive in long format -- one row per person per item.

```{r}
responses <- utils::read.csv(
  system.file("extdata", "example_responses.csv", package = "winstepsR"),
  colClasses = c("character", "character", "integer")
)
head(responses, 4)
nrow(responses)
```

The item difficulties were calibrated somewhere else. Here they are anchors:
values Winsteps is told to hold fixed rather than estimate.

```{r}
anchor_tbl <- utils::read.csv(
  system.file("extdata", "example_anchors.csv", package = "winstepsR"),
  colClasses = c("character", "numeric", "character")
)

anchors <- winsteps_anchors(anchor_tbl$item, anchor_tbl$value)
anchors
```

`winsteps_anchors()` exists because item names and anchor values are two vectors
that have to stay in the same order -- sequence number *is* position in the
files below. Keeping them in one object makes that structural rather than
something every call has to remember.

## The fixed-width layout

This is the part worth understanding, because it is where silent errors live.

Winsteps reads its data file by **column position**. Each line is a person: a
left-justified ID padded to a fixed width, a delimiter, then one character per
item.

```{r}
prepared <- winsteps_prepare_person_data(
  responses,
  id_col = "person_id", item_col = "item", score_col = "score",
  item_order = anchor_tbl$item
)
prepared
```

The printed layout is the whole contract. `NAMLEN=4` says the person name
occupies columns 1--4. `ITEM1=6` says the first response is at column 6.
`NI=12` says read twelve of them. If any of those three disagrees with the
actual bytes on the line, Winsteps does not complain -- it reads whatever is at
those positions and returns measures computed from the wrong responses.

Three consequences follow, and the package enforces all three:

- **Every response must be exactly one character.** A partial-credit score of
  `10` would occupy two columns and shift every item after it. That is an error
  here, not a silent shift.
- **Widths are counted in bytes, not characters.** A person ID containing an
  accented letter is more bytes than characters, so padding by character count
  would move the responses.
- **`item_order` decides the column order.** Supply it. Without it the order
  comes from whichever item appears first in the data, so re-sorting the input
  silently re-orders the columns.

Note the missing responses, written as `.`:

```{r}
grep(".", prepared$lines, fixed = TRUE, value = TRUE)[1:3]
```

`CODES=01` lists the valid responses, so anything else -- including `.` -- is
treated as not administered.

## The four files going in

`winsteps_estimate()` writes them all. With `run = FALSE` it stops before
invoking Winsteps, which is what lets this vignette build anywhere.

```{r}
run <- winsteps_estimate(
  data = responses,
  id_col = "person_id", item_col = "item", score_col = "score",
  anchors = anchors,
  run_id = "anatomy", working_dir = tempfile(),
  control_args = list(tfile = "17.1"),
  winsteps_exe = "C:/Winsteps/Winsteps.exe",
  run = FALSE
)
run
```

**The data file** is the layout above, one line per person:

```{r}
head(run$contents$data, 3)
```

**The anchor file** (`IAFILE=`) gives each item's sequence number and its fixed
value. The item name after the `;` is a comment -- Winsteps never reads it,
which matters in a moment:

```{r}
head(run$contents$anchor, 3)
```

**The control file** is the instructions. Everything the layout section
described appears here as a keyword:

```{r}
writeLines(run$contents$control)
```

Two things to notice. The item names appear again after `&END`, as *labels* --
that is the only way Winsteps learns them, because the names in the anchor file
are comments. Without those labels the item output comes back named `I0001`,
`I0002`, and cannot be joined to your items.

And `PFILE=` and `IFILE=` are requests: they name the files Winsteps should
write on the way out.

**The batch file** runs it:

```{r}
writeLines(run$contents$bat)
```

The first line makes the file self-locating, so it can be launched from any
directory and still find its control file. Paths are quoted for `cmd.exe` no
matter which platform generated them -- a file generated on macOS runs on
Windows unchanged.

## The files coming back

On Windows, `run = TRUE` would execute the batch file and read the results. The
files below are the real output from exactly this input.

**The person file** (`PFILE=`) is one row per person:

```{r}
measures <- winsteps_read_person_output(
  system.file("extdata", "example_full_person.out", package = "winstepsR")
)
measures[1:4, c("ENTRY", "NAME", "MEASURE", "COUNT", "SCORE", "MODLSE")]
```

`NAME` is deliberately read as text. Left to type guessing it would come back
as character for IDs like `00123` and numeric for IDs like `123` -- so a join on
that column would work until the day a cohort's IDs happened to lack leading
zeros.

**The item file** (`IFILE=`) is one row per item, and on an anchored run it is
how you check that the anchoring actually happened:

```{r}
items <- winsteps_read_item_output(
  system.file("extdata", "example_full_item.out", package = "winstepsR")
)
items[, c("ENTRY", "NAME", "MEASURE", "DISPL")]
```

`MEASURE` is the value that was supplied, unchanged. `DISPL` -- displacement --
is how far the data would have moved it. A run whose anchor file was ignored
completes normally and returns plausible measures; displacement is the evidence
that it was not ignored. What counts as too much displacement is a judgement
this package does not make for you.

```{r}
range(items$DISPL)
```

**The report file** holds any tables requested with `TFILE=`:

```{r}
report <- winsteps_read_report(
  system.file("extdata", "example_full_report.csv", package = "winstepsR")
)
report
summary(report)
```

Table layouts vary too much between tables to parse generically, so this stays
a character vector -- it just prints as a summary rather than as 215 lines.

## Reading the whole thing back

An existing control file can be parsed into the arguments that would write it,
which is the fastest way to bring an established Winsteps setup into R:

```{r}
winsteps_read_control_file(
  system.file("extdata", "example_full_control.ctr", package = "winstepsR")
)
```

## What each function writes or reads

| File | Winsteps keyword | Written by | Read by |
|---|---|---|---|
| Person data | `DATA=` | `winsteps_write_person_data()` | -- |
| Item anchors | `IAFILE=` | `winsteps_write_anchor_file()` | -- |
| Item subset | `IDFILE=` | `winsteps_write_item_subset_file()` | -- |
| Control | -- | `winsteps_write_control_file()` | `winsteps_read_control_file()` |
| Batch runner | -- | `winsteps_write_bat()` | -- |
| Person output | `PFILE=` | -- | `winsteps_read_person_output()` |
| Item output | `IFILE=` | -- | `winsteps_read_item_output()` |
| Table report | `TFILE=` | -- | `winsteps_read_report()` |

`winsteps_estimate()` is a wrapper that calls all of them in order. Every one
can be used on its own when the wrapper's assumptions do not fit.
