---
title: "Scoring a subset of items"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring a subset of items}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A common requirement is a measure for part of a test as well as the whole: a
content domain, a subscale, a reporting category. The item difficulties should
stay the same in both cases -- what changes is which items count.

Winsteps expresses this with `IDFILE=`, a list of item sequence numbers to
*exclude*. This package takes the inverse, since naming what you want is easier
than naming what you don't.

```{r setup}
library(winstepsR)

responses <- utils::read.csv(
  system.file("extdata", "example_responses.csv", package = "winstepsR"),
  colClasses = c("character", "character", "integer")
)
anchor_tbl <- utils::read.csv(
  system.file("extdata", "example_anchors.csv", package = "winstepsR"),
  colClasses = c("character", "numeric", "character")
)
anchors <- winsteps_anchors(anchor_tbl$item, anchor_tbl$value)
```

The example bank has twelve items split across two domains:

```{r}
anchor_tbl
```

## One domain

Pass the items to keep. Everything else in the bank is excluded:

```{r}
domain1 <- anchor_tbl$item[anchor_tbl$domain == "domain1"]
domain1

run <- winsteps_estimate(
  data = responses,
  id_col = "person_id", item_col = "item", score_col = "score",
  anchors = anchors,
  keep_items = domain1,
  run_id = "domain1",
  working_dir = tempfile(),
  winsteps_exe = "C:/Winsteps/Winsteps.exe",
  run = FALSE
)
run
```

The summary line reports both numbers -- twelve items anchored, six actually
estimated -- which is worth reading, because a domain filter that quietly
matched nothing is otherwise hard to notice.

The delete file is the inverse of what you asked for:

```{r}
run$contents$delete
```

Note that the **data file is unchanged**: all twelve responses are still
written, and the anchor file still carries all twelve values. Only the control
file's `IDFILE=` differs. That is deliberate -- the anchors stay identical
across every domain run, so measures from different domains are on the same
scale.

## What comes back

Real output from this domain run, shipped with the package:

```{r}
items <- winsteps_read_item_output(
  system.file("extdata", "example_domain1_item.out", package = "winstepsR")
)
items[, c("ENTRY", "NAME", "MEASURE", "DISPL")]
```

Two things worth knowing.

`ENTRY` keeps the sequence numbers from the **full** bank -- 1, 3, 5, 7, 9, 11 --
rather than renumbering the six retained items 1 to 6. So item output from a
subset run joins back to your item table on either `ENTRY` or `NAME`. Had
Winsteps renumbered, a join on `ENTRY` would have matched the wrong items
silently.

And `MEASURE` is the supplied anchor exactly, as it is in the full run. The
domain measures are on the same scale as everything else.

```{r}
persons <- winsteps_read_person_output(
  system.file("extdata", "example_domain1_person.out", package = "winstepsR")
)
persons[1:5, c("NAME", "MEASURE", "COUNT", "SCORE")]
```

`COUNT` is 6, not 12 -- each person was measured on the six domain items only.

## Every domain in one pass

Give each run a distinct `run_id` so their files do not collide:

```{r}
work <- tempfile()

domains <- split(anchor_tbl$item, anchor_tbl$domain)

by_domain <- Map(function(items, domain) {
  winsteps_estimate(
    data = responses,
    id_col = "person_id", item_col = "item", score_col = "score",
    anchors = anchors,
    keep_items = items,
    run_id = domain,
    working_dir = work,
    winsteps_exe = "C:/Winsteps/Winsteps.exe",
    run = FALSE
  )
}, domains, names(domains))

vapply(by_domain, function(r) basename(r$run_dir), character(1))
```

`run_id` names a subdirectory, so two runs sharing one would overwrite each
other. The paths are deterministic rather than unique -- that is useful, but it
means distinct names are your responsibility.

On Windows these can run in parallel: the batch file each run writes changes to
its own directory, so concurrent runs do not interfere.

## Choosing the domains

A caution that has nothing to do with the software. A domain made up of the
easiest items will have most candidates answering everything correctly, and
perfect scores carry no interior measure. The example data deliberately
alternates items between domains so both span the difficulty range:

```{r}
tapply(anchor_tbl$value, anchor_tbl$domain, range)
```

Whether your real domains do the same is a test-design question, but it shows up
here as a pile of extreme scores in the person output.
