---
title: "Getting started with winstepsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with winstepsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

winstepsR is the mechanical half of an R--Winsteps workflow: it writes the files
Winsteps needs, runs it, and reads the results back. It implements no scoring
rules of its own -- scale-score conversions, pass marks, domain definitions and
the like stay in your project, because they are specific to your program.

```{r setup}
library(winstepsR)
```

## Before you start

Winsteps runs on Windows only, and it is licensed separately. Point the package
at your executable once per session or project:

```{r, eval = FALSE}
options(winstepsR.exe_path = "C:/Winsteps/Winsteps.exe")
```

Everything except actually invoking Winsteps works on any platform, which is
what makes this vignette buildable. Pass `run = FALSE` to generate the input
files without running.

## What you supply

Two things: responses in long format, and the anchor values for your items.

```{r}
responses <- utils::read.csv(
  system.file("extdata", "example_responses.csv", package = "winstepsR"),
  colClasses = c("character", "character", "integer")
)
head(responses, 3)

anchor_tbl <- utils::read.csv(
  system.file("extdata", "example_anchors.csv", package = "winstepsR"),
  colClasses = c("character", "numeric", "character")
)
```

Pair the item names with their values. The order matters -- it is the order the
items will occupy in every file -- so keeping them together prevents them
drifting apart:

```{r}
anchors <- winsteps_anchors(anchor_tbl$item, anchor_tbl$value)
anchors
```

`winsteps_anchors()` validates once, here, rather than at each writer. A
duplicated item name, a missing anchor value or mismatched lengths are caught
now instead of producing a file Winsteps would read incorrectly.

## Estimating

```{r}
result <- winsteps_estimate(
  data = responses,
  id_col = "person_id", item_col = "item", score_col = "score",
  anchors = anchors,
  run_id = "spring_form_b",
  working_dir = tempfile(),
  winsteps_exe = "C:/Winsteps/Winsteps.exe",
  run = FALSE            # drop this on Windows to actually run
)
result
```

With `run = TRUE` the result also carries `$results` (person measures) and
`$items` (item measures). Here is what those look like, using real output from
this same data:

```{r}
measures <- winsteps_read_person_output(
  system.file("extdata", "example_full_person.out", package = "winstepsR")
)
measures[1:5, c("ENTRY", "NAME", "MEASURE", "COUNT", "SCORE")]
```

Column names are left exactly as Winsteps writes them. Renaming `NAME` to
whatever your project calls it is your business, not the package's.

## Checking the run before you trust it

Two habits worth forming.

**Check that the anchors were applied.** A run whose anchor file was ignored
finishes normally and returns plausible measures. The item output is the
evidence:

```{r}
items <- winsteps_read_item_output(
  system.file("extdata", "example_full_item.out", package = "winstepsR")
)
items[, c("NAME", "MEASURE", "DISPL")]
```

`MEASURE` should be the value you supplied. `DISPL` is how far the data would
have moved it -- large displacement means the anchors and the data disagree,
which is worth understanding before the measures are used.

**Look at the distribution.** `summary()` reports it and nothing more; no fit
flags, no cut-offs, because every cut-off is a decision belonging to your
program:

```{r}
c(mean = mean(measures$MEASURE), sd = stats::sd(measures$MEASURE))
table(measures$SCORE)
```

The full table stays in `$results`, so anything the summary leaves out you can
compute directly.

## Generating files on one machine, running on another

Because only the run step needs Windows, you can prepare everything anywhere:

```{r, eval = FALSE}
result <- winsteps_estimate(..., working_dir = "prepared_runs", run = FALSE)
```

Copy the run directory to a Windows machine and double-click `run.bat`, or call
`winsteps_run()` there. The batch file changes to its own directory first, so it
works wherever it is put.

`result$contents` holds the lines of every file that was written, so a run can
be inspected afterwards even when `working_dir` was a temporary directory the
operating system has since cleared:

```{r}
names(result$contents)
head(result$contents$data, 2)
head(result$contents$anchor, 2)
```

## When the wrapper does not fit

`winsteps_estimate()` composes eight smaller functions, each usable alone:
`winsteps_prepare_person_data()`, `winsteps_write_person_data()`,
`winsteps_write_anchor_file()`, `winsteps_write_item_subset_file()`,
`winsteps_write_control_file()`, `winsteps_write_bat()`, `winsteps_run()`, and
the readers. See `vignette("anatomy-of-a-run")` for what each file contains and
why the layout matters.
